miniKanren with fair search

### 2019-03-22 Paper/paper.tex
80ch per line

The syntax of a programming language should reflect its semantics.
When using a disjunction operator in relational programming, a programmer would
expect all clauses of this disjunction to share the same chance of being explored,
as these clauses are written in parallel. 
% The existing disjunctive operators in miniKanren, however, prioritize their clauses
The existing multiarity disjunctive operator in miniKanren, however, prioritize its clauses
by the order of which these clauses are written down. 
% We have devised a new search strategy that searches evenly in all clauses.
We have devised two new search strategies that allocate computational effort fairly in all clauses.

@@

@@

With interleaving depth-first search, disjunctive operator allocates half computational effort to its first goal, and quater to the second, eighth to the third, and so on. Consequently, miniKanren programmers sometimes need to order their \texttt{conde} clauses carefully to prevent some clauses absorbing unproportional computational effort. For example, when a relational proof checker is used as a proof generator, the eliminational rules should comes later.
%TODO KC: prefer an example from other literatures 
Seasoned miniKanreners usually know how to make use of this unfair allocation to optimize their programs. However, we believe search strategies less sensitive to goal order can also be useful to little miniKanreners as well as seansoned ones. We propose two such search strategies, balanced interleaving DFS (biDFS) and breadth-first search (BFS), and observe how they affect the efficiency and query result of known miniKanren programs. The experiment is conducted with the miniKanren from \textit{The Reasoned Schemer, 2nd Edition}.

@@

(TODO: summarize following sections) Section 2 explains why iDFS biases toward left \texttt{conde} clauses. Section 3 is about the biDFS. Section 3 is about BFS.

@@

\texttt{conde} clauses are combined by \texttt{disj2}, right associatively. The core functionality of \texttt{disj2} is completed by \texttt{append-inf} (Fig.~\ref{disj2-and-append-inf}). When both input streams are infinite, the resulting mature part contains only the mature part of \texttt{s-inf}. The whole \texttt{t-inf} goes to the resulting immature part. However, \texttt{t-inf} and \texttt{s-inf} are swapped in the delayed recursive call. Hence the search strategy spend computational effort evenly in two appended streams. As \texttt{disj2} is right associative, the left clauses are nested in less calls to \texttt{append-inf}, which implies that they are more likeyly being computed.

@@

For the above \texttt{run}, both search strategies produces answers in increasing
order of costs, i.e. both of them are \emph{cost-respecting}. In more complicated
cases, however, interleaving DFS might not produce answers in cost-repecting order.
For instance, with iDFS the \texttt{run} in Fig.~\ref{conde-repeato-iDFS} produces
answers in a seemingly random order. In contrast, the same run with BFS produces
answers in an expected order (Fig.~\ref{conde-repeato-BFS}).

@@

Streams are cost respective when they are initially constructed by \texttt{==}. However, the \texttt{mk-0} version of \texttt{append-inf} (Fig.~\ref{append-inf-0}) breaks cost respectiveness if its first input stream, \texttt{s-inf}, is infinite. The resulting mature part contains only the mature part of \texttt{s-inf}. The whole \texttt{t-inf} goes to the resulting immature part.

@@

The \texttt{mk-1} version of \texttt{append-inf} (Fig.~\ref{append-inf-1}) restores cost-respectiveness by combining the mature parts in the fashion of \texttt{append}. This \texttt{append-inf} calls its helper immediately, with the first argument, \texttt{s?}, set to \texttt{\#{}t}, which means \texttt{s-inf} in the helper is the \texttt{s-inf} in the driver. Two streams are swapped in the third \texttt{cond} clause, with \texttt{s?} flipped accordingly.

@@

\texttt{mk-1} is not efficient in two aspects. \texttt{append-inf} need to copy all \texttt{cons} cells of two input streams when the first stream has a non-trivial immature part. Besides, \texttt{mk-1} computes answers of the same cost at once, even when only a portion is queried. We solves the two problems in the next subsections.

@@

The last interesting function is \texttt{take-inf} (Fig.~\ref{take-inf-3-1}). The parameter \texttt{vs} is a list of solutions. The next two parameters together represents a functional queue in a typical way. The first two \texttt{cond} lines are very similar to their counterparts in mk-0 and mk-1. The third line runs when we exaust all solutions. The forth line re-shape the queue. The fifth and last line invoke the first thunk in the queue and use the mature part of the resulting stream, \texttt{s-inf}, as the new \texttt{vs}, and enqueuing \texttt{s-inf}'s thunks.

>>>>

The syntax of a programming language should reflect its semantics. When using a
disjunction operator in relational programming, a programmer would expect all
clauses of this disjunction to share the same chance of being explored, as
these clauses are written in parallel. The existing multiarity disjunctive
operator in miniKanren, however, prioritize its clauses by the order of which
these clauses are written down. We have devised two new search strategies that
allocate computational effort fairly in all clauses.

@@

%TODO 1. reformat LaTeX (80 chars per line)
%TODO 2. stat quine
%TODO 3. keep writing

@@

With interleaving depth-first search, disjunctive operator allocates half
computational effort to its first goal, and quater to the second, eighth to the
third, and so on. If a goal that can generate planty of unwanted partial 
solutions is placed at the beginning of a \texttt{conde} expressions, it might
waste a lot computational resource. Consequently, miniKanren programmers 
sometimes need to order their \texttt{conde} clauses carefully to prevent some 
clauses absorbing unproportional computational effort. For example, when a 
relational proof checker is used as a proof generator, the eliminational rules 
should comes later.
%TODO KC: prefer an example from other literatures
Seasoned miniKanreners usually know how to utilize the unfairness to optimize
their programs. However, we believe search strategies less sensitive to goal
order can also be useful to little miniKanreners as well as seansoned ones. We 
propose two such search strategies, balanced interleaving DFS (biDFS) and
breadth-first search (BFS), and observe how they affect the efficiency and
query result of known miniKanren programs. The experiment is conducted with the
miniKanren from \textit{The Reasoned Schemer, 2nd Edition}.

@@

(TODO: summarize following sections) Section 2 explains why iDFS biases toward left \texttt{conde} clauses. Section 3 is about the biDFS. Section 4 is about BFS.

@@

\texttt{conde} clauses are combined by \texttt{disj2}, right associatively. The
core functionality of \texttt{disj2} is completed by \texttt{append-inf} (Fig.~
ref{disj2-and-append-inf}). When both input streams are infinite, the resulting
mature part contains only the mature part of \texttt{s-inf}. The whole
\texttt{t-inf} goes to the resulting immature part. However, \texttt{t-inf} and
\texttt{s-inf} are swapped in the delayed recursive call. Hence the search
strategy spend computational effort evenly in two appended streams. As
\texttt{disj2} is right associative, the left clauses are nested in less calls
to \texttt{append-inf}, which implies that they are more likeyly being
computed.

@@

For the above \texttt{run}, both search strategies produces answers in 
increasing order of costs, i.e. both of them are \emph{cost-respecting}. In 
more complicated cases, however, interleaving DFS might not produce answers in 
cost-repecting order. For instance, with iDFS the \texttt{run} in Fig.~
\ref{conde-repeato-iDFS} produces answers in a seemingly random order. In 
contrast, the same run with BFS produces answers in an expected order (Fig.~
\ref{conde-repeato-BFS}).

@@

Streams are cost respective when they are initially constructed by \texttt{==}. 
However, the \texttt{mk-0} version of \texttt{append-inf} (Fig.~\ref{append-
inf-0}) breaks cost respectiveness if its first input stream, \texttt{s-inf}, 
is infinite. The resulting mature part contains only the mature part of 
\texttt{s-inf}. The whole \texttt{t-inf} goes to the resulting immature part.

@@

The \texttt{mk-1} version of \texttt{append-inf} (Fig.~\ref{append-inf-1}) 
restores cost-respectiveness by combining the mature parts in the fashion of 
\texttt{append}. This \texttt{append-inf} calls its helper immediately, with 
the first argument, \texttt{s?}, set to \texttt{\#{}t}, which means \texttt{s-
inf} in the helper is the \texttt{s-inf} in the driver. Two streams are swapped 
in the third \texttt{cond} clause, with \texttt{s?} flipped accordingly.

@@

\texttt{mk-1} is not efficient in two aspects. \texttt{append-inf} need to copy 
all \texttt{cons} cells of two input streams when the first stream has a non-
trivial immature part. Besides, \texttt{mk-1} computes answers of the same cost 
at once, even when only a portion is queried. We solves the two problems in the 
next subsections.

@@

The last interesting function is \texttt{take-inf} (Fig.~\ref{take-inf-3-1}). 
The parameter \texttt{vs} is a list of solutions. The next two parameters, 
\texttt{P} and \texttt{Q}, together represent a queue. The first two 
\texttt{cond} lines are very similar to their counterparts in mk-0 and mk-1. 
The third line runs when we exaust all solutions. The forth line re-shape the 
queue. The fifth and last line invoke the first thunk in the queue and use the 
mature part of the resulting stream, \texttt{s-inf}, as the new \texttt{vs}, 
and enqueuing \texttt{s-inf}'s thunks.
